Quasi-Monte Carlo Methods: What, Why, and How?

### 2025-08-05 HickernellMCQMC2024_QMCTutorial_rev_3.tex
aleksei: put all updates in rev3

Another family of LD sequences that is a generalization of the van der Corput sequence \eqref{eq:vdc} is digital sequences \cite{DicPil10a,Nie92}.  For simplicity, we restrict ourselves to $b = 2$ and let $\oplus$ denote binary digitwise addition modulo $2$ (also known as digitwise exclusive or), e.g., $3/8 \oplus 3/4 = {}_20.011 \oplus {}_20.110 = {}_20.101 = 5/8$.  A digital sequence is defined as
\begin{equation} \label{eq:digital}
	\bsx_i^{\text{dig}} := i_0 \bsx^{\text{dig}} _1 \oplus i_1 \bsx^{\text{dig}}_2 \oplus i_2 \bsx^{\text{dig}} _4 \oplus \cdots \in [0,1)^d \quad \text{for }
	i = i_0 + i_12 + i_2 2^2 + \cdots,
\end{equation}
where $\bsx^{\text{dig}}_1, \bsx^{\text{dig}}_2, \bsx^{\text{dig}}_4, \bsx^{\text{dig}}_8, \ldots$ are carefully chosen to ensure low discrepancy.  (For $d=1$ and $x^{\text{dig}}_{2^m} = 2^{-m-1}$, this is the van  der Corput sequence in base $2$.) The node set $\{ \bsx^{\text{dig}}_0, \ldots, \bsx^{\text{dig}}_{2^m-1}\}$  for non-negative integer $m$ is called a digital net and is closed under $\oplus$.

@@

Figure \ref{fig:scaledweightedcentered} (bottom) plots the scaled centered discrepancy with coordinate weights $\bsgamma = (1/\ell)_{\ell =1}^d$, i.e., 
\begin{equation*}
    \text{wt-ctr-discrepancy}\bigl(\{ \bsx_i\}_{i=0}^{n-1},\bsgamma \bigr)/\text{wt-ctr-discrepancy}(\emptyset,\bsgamma ),
\end{equation*}
against sample size for a variety of $d$.  Now, even for larger $d$ there is a reasonable decay that approximates $\mathcal{O}(n^{-1})$.

@@

One idea for constructing confidence intervals without replications is to approximate the decay rate of Fourier coefficients of $f$. Bounds of absolute certainty have be derived for cones of functions with reasonable decay of their Fourier coefficients \cite{HicEtal17a}. As in the qMC-CLT stopping criterion, one may iteratively double $n$ until the bounds are sufficiently tight. Figure \ref{fig:asian_option_pda_vs_chol} compares these decay-tracking qMC algorithms against IID Monte Carlo stopping criterion across a variety of tolerances for pricing an Asian option from finance \cite[p.\ 8]{Gla03}.

@@

\begin{figure}
	\centering
 %    \includegraphics[width = 0.48\textwidth]{\figpath//NumSamples_Asian_Option_tol_0.001_d52.pdf}
	% \includegraphics[width = 0.48\textwidth]{\figpath/Times_Asian_Option_tol_0.001_d52.pdf} \quad 
    \includegraphics[width = 1\textwidth]{\figpath//Times_NumSamples_Asian_Option_tol_0.001_d52.pdf}
	\caption{Comparison of IID Monte Carlo and qMC stopping criteria for estimating the fair price of an Asian option with arithmetic mean. For QMC with Sobol' points, using a principal component analysis (PCA) decomposition of the Brownian motion provides substantial savings compared to the Cholesky decomposition. Using an Asian option with geometric mean control variate (CV) provides additional savings. \label{fig:asian_option_pda_vs_chol}}
\end{figure}

>>>>

Another family of LD sequences that generalize the van der Corput sequence \eqref{eq:vdc} are digital sequences \cite{DicPil10a,Nie92}.  For simplicity, we restrict ourselves to $b = 2$ and let $\oplus$ denote binary digitwise addition modulo $2$ (also known as digitwise exclusive or), e.g., $3/8 \oplus 3/4 = {}_20.011 \oplus {}_20.110 = {}_20.101 = 5/8$.  A digital sequence is defined as
\begin{equation} \label{eq:digital}
	\bsx_i^{\text{dig}} := i_0 \bsx^{\text{dig}} _1 \oplus i_1 \bsx^{\text{dig}}_2 \oplus i_2 \bsx^{\text{dig}} _4 \oplus \cdots \in [0,1)^d \quad \text{for }
	i = i_0 + i_12 + i_2 2^2 + \cdots,
\end{equation}
where $\bsx^{\text{dig}}_1, \bsx^{\text{dig}}_2, \bsx^{\text{dig}}_4, \bsx^{\text{dig}}_8, \ldots$ are carefully chosen to ensure low discrepancy.  (For $d=1$ and $x^{\text{dig}}_{2^m} = 2^{-m-1}$, this is the van  der Corput sequence in base $2$.) The node set $\{ \bsx^{\text{dig}}_0, \ldots, \bsx^{\text{dig}}_{2^m-1}\}$  for non-negative integer $m$ is called a digital net and is closed under $\oplus$.

@@

Figure \ref{fig:scaledweightedcentered} (bottom) plots the scaled centered discrepancy with coordinate weights $\bsgamma = (1/\ell)_{\ell =1}^d$, i.e., 
\begin{equation*}
    \text{wt-ctr-discrepancy}\bigl(\{ \bsx_i\}_{i=0}^{n-1},\bsgamma \bigr)/\text{wt-ctr-discrepancy}(\emptyset,\bsgamma ),
\end{equation*}
against sample size for a range of $d$.  Now, even for larger $d$ there is a reasonable decay that approximates $\mathcal{O}(n^{-1})$.

@@

One idea for constructing confidence intervals without replications is to approximate the decay rate of Fourier coefficients of $f$. Bounds of absolute certainty have be derived for cones of functions with reasonable decay of their Fourier coefficients \cite{HicEtal17a}. As in the qMC-CLT stopping criterion, one may iteratively double $n$ until the bounds are sufficiently tight. Figure \ref{fig:asian_option_pda_vs_chol} compares these decay-tracking qMC algorithms against IID Monte Carlo stopping criterion across a range of tolerances for pricing an Asian option from finance \cite[p.\ 8]{Gla03}.

@@

\begin{figure}
	\centering
 %    \includegraphics[width = 0.48\textwidth]{\figpath//NumSamples_Asian_Option_tol_0.001_d52.pdf}
	% \includegraphics[width = 0.48\textwidth]{\figpath/Times_Asian_Option_tol_0.001_d52.pdf} \quad 
    \includegraphics[width = 1\textwidth]{\figpath//Times_NumSamples_Asian_Option_tol_0.001_d52.pdf}
	\caption{Comparison of IID Monte Carlo and qMC stopping criteria for estimating the fair price of an Asian option with arithmetic mean. For qMC with Sobol' points, using a principal component analysis (PCA) decomposition of the Brownian motion provides substantial savings compared to the Cholesky decomposition. Using an Asian option with geometric mean control variate (CV) provides additional savings. \label{fig:asian_option_pda_vs_chol}}
\end{figure}
